A Survey of Quantitative Descriptions of Molecular Structure

### 2012-07-29 paper.tex
final set of edits

The requirements listed here ensures that subsequent analyses
involving descriptors are robust. But in addition to numerical
robustness, it is important for many scenarios that the descriptor
have some form of physical interpretability.  For example, a model
that employs descriptors to predict cytotoxicity will be most useful if
it is numerically accurate as well as providing insight into which
structural features confer toxicity. In such a scenario, abstract
descriptors do not allow us to provide a physical explanation of the
models predictions. Another scenario is the ability to map descriptor
values back to the structure for visualization purposes such as the
``Glowing Molecule'' depictions of Segall et al
\cite{Segall:2009fk}. Such visualizations are practical only when
descriptor values can be related to structural features. Examples
include surface based descriptors such as CPSA's as well as explicit
structural descriptors such as paths, chains etc.

@@

@@

\ctable[caption={A comparison of the VABC descriptor for molecular
  volume (as implemented by the CDK) and other approximate
  implementations of molecular volume (as implemented by MOE
  2008.10). All values are in \AA$^{3}$ and both plots are overlaid with a loess fit.},cap={},label={fig:molvol},botcap,figure]
{c}
{}
{
  \includegraphics[width=\linewidth]{vol-moe-cdk}
}

>>>>

The requirements listed here ensures that subsequent analyses
involving descriptors are robust. But there are other aspects of a
descriptor that can make it more or less useful. For example,a
descriptor that can be rapidly calculated and is not dependent on
experimental properties can be considered more useful than one that is
computationally intensive and / or dependent on experimental
results. In addition, it is important for many scenarios that the descriptor
have some form of physical interpretability.  For example, a model
that employs descriptors to predict cytotoxicity will be most useful if
it is numerically accurate as well as providing insight into which
structural features confer toxicity. In such a scenario, abstract
descriptors do not allow us to provide a physical explanation of the
models predictions. Another scenario is the ability to map descriptor
values back to the structure for visualization purposes such as the
``Glowing Molecule'' depictions of Segall et al
\cite{Segall:2009fk}. Such visualizations are practical only when
descriptor values can be related to structural features. Examples
include surface based descriptors such as CPSA's as well as explicit  
structural descriptors such as paths, chains etc.

@@

This discussion has focused on aspects of descriptors that make them
useful in computational models. It is also relevant to note that the
utility of a descriptor can be dependent on the problem being
considered. More specifically, certain end points may require specific
molecular features to be taken into account. Descriptors that do not,
will not be useful in developing models for that endpoint. For
example, an accurate boiling point model will require descriptors that
capture information regarding molecular branching, hydrogen bonding
potential and so on. Similarly, a model to predict activity against a
cytoplasmic or nuclear target, will likely require that molecular
characteristics enabling passage across the cell membrane be taken
into account. In such a case, a descriptor characterizing the
lipophilicity of the molecule will be useful. These points also apply
to other types of computational approaches, such as clustering. In the
end, the utility of a descriptor is determined by whether it encapsulates
sufficient and relevant information. In general a single descriptor
will not satisfy these requirements and thus one usually works with
sets of descriptors.

@@

\ctable[caption={A comparison of the VABC descriptor for molecular
  volume (as implemented by the CDK) and other approximate
  implementations of molecular volume (a 2D approach and a grid based
  approach), as implemented by MOE
  2008.10. All values are in \AA$^{3}$ and both plots are overlaid with a loess fit.},cap={},label={fig:molvol},botcap,figure]
{c}
{}
{
  \includegraphics[width=\linewidth]{vol-moe-cdk}
}
